### 2025-05-13 paper/paper.qmd
very rough draft of paper

---
title: "Respiratory Disease Outbreaks in Toronto Pre, During, and Post Covid-19 Pandemic "
subtitle: How Coronavirus and Other Respiratory Diseases have responded to the covid-19 pandemic and Toronto pandemic policies
author: 
  - Evelyn Hughes
thanks: "Code and data are available at: [https://github.com/EveHughes/Donaldson-Paper)."
date: 05/12/2025
date-format: long

abstract: "Through the analysis of Outbreaks in Toronto Healthcare Institutions, this paper seeks to identify the trends in respiratory infections in Toronto during the implementation and later lifting of covid-19 preventatory policies, including general trends in number of outbreaks, and makeup of respiratory infections. Through data visualization, this analysis finds that while reported coronavirus outbreaks have lowered since pandemic times, “secondary” respiratory diseases such as Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus that were at lower levels during the pandemic have begun to rise again. At the same time, covid becomes a lesser percentage of overall respiratory diseases. Given different respiratory diseases require different treatment, these findings are relevant for informing healthcare facilities to place a lesser emphasis on covid-19, etc."

format: pdf
number-sections: true
bibliography: references.bib
---

```{python}
#| label: fig-polar
#| fig-cap: "A line plot on a polar axis"

import numpy as np
import matplotlib.pyplot as plt

r = np.arange(0, 2, 0.01)
theta = 2 * np.pi * r
fig, ax = plt.subplots(
  subplot_kw = {'projection': 'polar'} 
)
ax.plot(theta, r)
ax.set_rticks([0.5, 1, 1.5, 2])
ax.grid(True)
plt.show()
```

# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

\newpage

# References

### 2025-05-13 paper/paper.qmd
added very rough draft of paper

@@
 ```{python}
-#| label: fig-polar
+#| label: fig-polar-1
 #| fig-cap: "A line plot on a polar axis"
@@
 
-Overview paragraph
+On January 25, 2020, the first case of COVID-19 was reported in Canada, in a healthcare facility in Toronto (Cheese 2025).
+Around two months later on March 17, Doug Ford declared a state of emergency for Ontario, and the closure of facilities such as schools.
+Additionally, mask mandates, social distancing guidelines, and vaccine cards were all put into place to mitigate the spread of the disease.
+Later around 2022, these restrictions began to lift.
+Schools went back in person, etc, etc, etc.
+(Compartin 2022).
+As the restrictions lift, it becomes important to characterize the effect they hold upon respiratory diseases both as a whole, and specific diseases.
 
-Estimand paragraph
+Previously, in 2021, research was conducted to examine the effects of covid protocol on other respiratory diseases, finding evidence for a decrease in disease during covid protocol (Groves et al 2021).
+However, there is a lack of literature examining the effects of the lifting of protocol on these same diseases, and examining the general trends.
+This research aims to fill that gap, and identify how diseases have reacted to the lifting of covid protocol in relation to both their pre covid (2019), and during covid (2020-2022 ish) levels.
+Doing such can inform policy that encourages disease-specific preventative measures, and promotion of vaccines to prevent increased outbreaks.
 
-Results paragraph
+This analysis first examines the trends in the overall number of outbreaks and respiratory outbreaks each year to better understand how the number of outbreaks has been varying over the years and identify general trends.
+Then, it looks at coronavirus, Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus.
+Finally, it examines the trends in percentage of respiratory outbreaks with covid as an agent.
+Through the data visualization, it was found that while coronavirus outbreaks have been falling since their covid levels, other respiratory diseases have been seeing a subsequent increase in cases post pandemic.
+Additionally, the percentage of respiratory outbreaks with covid as an agent has been decreasing since the pandemic.
+So, while an emphasis on covid specific protocols is important, preventative measures for other respiratory diseases, such as RSV vaccines, that were overshadowed during the pandemic should be further encouraged (Public Health Agency of Canada 2024).
 
-Why it matters paragraph
-
-Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....
+In the sections that follow are a description of the data, visualization of the data and discussion of the results.
 
@@
 
-## Overview
+## Raw Data
 
-We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...
+Outbreaks in Toronto Healthcare Institutions data is provided by Toronto Public Health., and is obtained from the OpenData Toronto Portal.
+The dataset is updated weekly on Thursdays with data as of the Wednesday at 2pm.(https://open.toronto.ca/dataset/outbreaks-in-toronto-healthcare-institutions/ ).
 
-Overview text
+The dataset consists of data from years 2016-2025 inclusive, with each containing instances of outbreaks in toronto healthcare institutions.
 
-## Measurement
+“Under the Ontario Health Protection and Promotion Act (HPPA), healthcare institutions (hospitals, long-term care homes and retirement homes) are required to monitor staff and patients/residents for signs and symptoms of gastroenteric (e.g., nausea, vomiting, diarrhea, fever) and respiratory (e.g., cough, runny nose, sore throat, fever) infections.
+Healthcare institutions must also actively look for, detect and report suspected and/or confirmed outbreaks to their local public health unit”
 
-Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.
+For every record there is an id number, institution name, institution address, outbreak setting (type of facility), type of outbreak, causative agent-1, causative agent-2, date outbreak began, date outbreak ended, and whether the outbreak is currently active.
 
-## Outcome variables
+The type of outbreak is listed as Respiratory, Enteric, or Other.
+By the Ministry of Health, a respiratory outbreak is constituted by “Two cases of acute respiratory infections (ARI) within 48 hours with any common epidemiological link (e.g., unit, floor), at least one of which must be laboratory-confirmed; OR Three cases of ARI (laboratory confirmation not necessary) occurring within 48 hours with any common epidemiological link (e.g., unit, floor).
+https://www.toronto.ca/wp-content/uploads/2023/06/9942-Outbreak-Preparedness-ToolkitRespiratoryTableTopFacilitators.pdf
 
-Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.
+Causative Agent-1 is listed as either as the specific causative agent (eg Covid-19), or is listed as "Unable to Identify". 
+Causative Agent-2 is listed as either the specific secondary causative agent, or is listed as "None".
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+Below is a preview of this information:
+
+```{python}
+#| label: fig-polar
+#| fig-cap: "A line plot on a polar axis"
+
+import numpy as np
+import matplotlib.pyplot as plt
+
+r = np.arange(0, 2, 0.01)
+theta = 2 * np.pi * r
+fig, ax = plt.subplots(
+  subplot_kw = {'projection': 'polar'} 
+)
+ax.plot(theta, r)
+ax.set_rticks([0.5, 1, 1.5, 2])
+ax.grid(True)
+plt.show()
+```
+
+For this analysis, the variables of interest are the yearly number of outbreaks for Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus over the years of 2019 - 2024 inclusive.
+
+## Analysis Data
+
+For each year, the number of cases were aggregated, and were aggregated within each respective category.
+The diseases of interest were: Coronavirus Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus.
+Instances of coronavirus and covid-19 were grouped together under the title of “coronavirus”. 
+Influenza A & B were additionally grouped together under the title “Influenza”.
+
+Below is a sample of the cleaned data:
 
@@
 
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+And the summary statistics for the cleaned data:
 
@@
 
-Talk way more about it.
-
-## Predictor variables
-
-Add graphs, tables and text.
-
-Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.
-
-# Model
-
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
-
-## Model set-up
-
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
-
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
-
-### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
-
 # Results
 
-Our results are summarized in @tbl-modelresults.
-
+## Total Outbreak, Coronavirus Outbreak, and Respiratory Outbreak 
 ```{r}
@@
 ```
+
+## Influenza Outbreak & Other
 
@@
 
+## Other outbreaks
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-modelresults-2
+#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
+#| warning: false
+
+modelsummary::modelsummary(
+  list(
+    "First model" = first_model
+  ),
+  statistic = "mad",
+  fmt = 2
+)
+```
+
 # Discussion
 
-## First discussion point {#sec-first-point}
+## Total Outbreak, Coronavirus Outbreak, and Respiratory Outbreak {#sec-first-point}
 
@@
 
-## Second discussion point
+## Influenza Outbreak & Other
 
@@
 
-## Third discussion point
+## Other outbreaks
 
@@
 
-Weaknesses and next steps should also be included.
+One main weakness with this analysis is it is limited to healthcare facilities, which fail to be representative of the larger demographic of Toronto.
+A next step would be to analyze disease outbreak in the general population.
 
 \newpage
-
-\appendix
-
-# Appendix {-}
 

### 2025-05-13 paper/paper.qmd
previewed rough rough draft

@@
 ```{python}
-#| label: fig-polar-1
+#| label: fig-polar
 #| fig-cap: "A line plot on a polar axis"
@@
 For every record there is an id number, institution name, institution address, outbreak setting (type of facility), type of outbreak, causative agent-1, causative agent-2, date outbreak began, date outbreak ended, and whether the outbreak is currently active.
+Below is a preview of this information:
 
@@
 
-Causative Agent-1 is listed as either as the specific causative agent (eg Covid-19), or is listed as "Unable to Identify". 
-Causative Agent-2 is listed as either the specific secondary causative agent, or is listed as "None".
+Causative Agent-1 is either a specific agent (ie: Coronavirus), or "Unable to identify".
+Causative Agent-2 is either a specific agent, or "None".
 
-Below is a preview of this information:
+*insert table showing first few lines of raw data
 
-```{python}
-#| label: fig-polar
-#| fig-cap: "A line plot on a polar axis"
-
-import numpy as np
-import matplotlib.pyplot as plt
-
-r = np.arange(0, 2, 0.01)
-theta = 2 * np.pi * r
-fig, ax = plt.subplots(
-  subplot_kw = {'projection': 'polar'} 
-)
-ax.plot(theta, r)
-ax.set_rticks([0.5, 1, 1.5, 2])
-ax.grid(True)
-plt.show()
-```
-
-For this analysis, the variables of interest are the yearly number of outbreaks for Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus over the years of 2019 - 2024 inclusive.
+For this analysis, the variables of interest are the yearly number of outbreaks for Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus.
+Over 2019-2024.
 
@@
 The diseases of interest were: Coronavirus Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus.
-Instances of coronavirus and covid-19 were grouped together under the title of “coronavirus”. 
+Instances of coronavirus and covid-19 were grouped together under the title of “coronavirus”, given covid is caused by a coronavirus.
 Influenza A & B were additionally grouped together under the title “Influenza”.
@@
 
-```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
-#| echo: false
+*** insert table
 
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
-```
+Below are the summary statistics for the cleaned data:
 
-And the summary statistics for the cleaned data:
-
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-
-analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))
-
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
-```
+*** insert summary statistics
 
@@
 
-## Total Outbreak, Coronavirus Outbreak, and Respiratory Outbreak 
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
+Total & respiratory & Coronavirus:
+*** insert graph
 
-library(rstanarm)
+We see that the total number of cases fluctuates with the number of respiratory cases, with number of outbreaks increasing since 2019 to 2022, and beginning to decrease afterwards
 
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
-```
+The number of cases also fluctuates with the number of respiratory cases and total cases, following the same trend described above.
 
-## Influenza Outbreak & Other
+Composition of Respiratory Illness:
+*** insert graph 
+Covid / Secondary Disease / Other & Unknown
 
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
+Influenza:
+Number of cases fall from 2019- 2021, with 0 cases in 2021, and increase thereafter.
 
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
-
-## Other outbreaks
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults-2
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
+Parainfluenza, RSV, Metapneumovirus, and Rhinovirus:
+Each has specifics, see general trend of lower number & some w/ slight decrease reported in pre-pandemic & pandemic, with a subsequent increase post.
 
@@
 
-## Total Outbreak, Coronavirus Outbreak, and Respiratory Outbreak {#sec-first-point}
+## Rise in Non-Covid Respiratory Outbreaks {#sec-first-point}
 
@@
 
-## Influenza Outbreak & Other
+## Decrease in Covid Percentage of Respiratory Outbreaks
 
@@
 
-## Other outbreaks
-
 ## Weaknesses and next steps
 
-One main weakness with this analysis is it is limited to healthcare facilities, which fail to be representative of the larger demographic of Toronto.
-A next step would be to analyze disease outbreak in the general population.
+Weaknesses and next steps should also be included.
 
 \newpage
+
+\appendix
 

### 2025-05-13 paper/paper.qmd
added rough code for graphs

@@
 ---
-
-```{python}
-#| label: fig-polar
-#| fig-cap: "A line plot on a polar axis"
-
-import numpy as np
-import matplotlib.pyplot as plt
-
-r = np.arange(0, 2, 0.01)
-theta = 2 * np.pi * r
-fig, ax = plt.subplots(
-  subplot_kw = {'projection': 'polar'} 
-)
-ax.plot(theta, r)
-ax.set_rticks([0.5, 1, 1.5, 2])
-ax.grid(True)
-plt.show()
-```
 
@@
 
-*insert table showing first few lines of raw data
+Below is an example of the raw data from 2019:
+
+```{python}
+#| label: fig-polar
+#| fig-cap: "A line plot on a polar axis"
+
+import polars as pl
+import matplotlib.pyplot as plt
+
+raw_data = pl.read_csv("Donalson-Paper/data/01-raw_data/ob_report_" + "2019" + ".csv")
+raw_data = raw_data.head(5)
+example_data = raw_data.to_dict(as_series = False)
+plt.table(example_data)
+
+plt.show()
+```
 
@@
 
-*** insert table
+```{python}
+#| label: disease-count-preview
+#| fig-cap: "A line plot on a polar axis"
 
-Below are the summary statistics for the cleaned data:
+import polars as pl
+import matplotlib.pyplot as plt
 
-*** insert summary statistics
+raw_data = pl.read_csv("Donalson-Paper/data/02-analysis_data/disease_count.csv")
+raw_data = raw_data.head(5)
+example_data = raw_data.to_dict(as_series = False)
+plt.table(example_data)
+
+plt.show()
+```
+```{python}
+#| label: yearly-count-preview
+#| fig-cap: "A line plot on a polar axis"
+
+import polars as pl
+import matplotlib.pyplot as plt
+
+raw_data = pl.read_csv("Donalson-Paper/data/02-analysis_data/yearly_disease_count.csv")
+raw_data = raw_data.head(5)
+example_data = raw_data.to_dict(as_series = False)
+plt.table(example_data)
+
+plt.show()
+```
+
+Below are the summary statistics for the cleaned yearly count data:
+
+```{python}
+#| label: yearly-count-summary-statistics
+#| fig-cap: "A line plot on a polar axis"
+
+import polars as pl
+import matplotlib.pyplot as plt
+
+raw_data = pl.read_csv("Donalson-Paper/data/02-analysis_data/yearly_disease_count.csv")
+summary_data = raw_data.describe()
+summary_data = summary_data.to_dict(as_series = False)
+plt.table(summary_data)
+
+plt.show()
+```
 
@@
 Total & respiratory & Coronavirus:
-*** insert graph
+```{python}
+#| label: corona-resp-total-graph
+#| fig-cap: "A line plot on a polar axis"
+
+import polars as pl
+import matplotlib.pyplot as plt
+import np as np
+
+t = np.arange(2019, 2025, 1)
+KEYS = ["Coronavirus", "Respiratory", "Total"]
+raw_data = pl.read_csv("Donalson-Paper/data/02-analysis_data/yearly_disease_count.csv")
+for key in KEYS:
+  plt.plot(t, raw_data[key], label = key, {'marker': 'o'})
+
+plt.set_title("Coronavirus, Respiratory & Total Outbreaks")
+plt.set_xlabel("Year")
+plt.set_ylabel("Outbreak Count")
+plt.grid()
+
+plt.legend()
+plt.show()
+```
 

### 2025-05-13 scripts/05-exploratory_data_analysis.py

@@ -7,48 +7,67 @@
 # Pre-requisites: [...UPDATE THIS...]
 # Any other information needed? [...UPDATE THIS...]
 
-
-#### Workspace setup ####
-import numpy as np
 import polars as pl
 import matplotlib.pyplot as plt
+import numpy as np
 
-#### Read data ####
-composition_data = pl.read_csv("data/02-analysis_data/disease_count.csv")
-dict1 = composition_data.to_dict(as_series = False)
-analysis_data = pl.read_csv("data/02-analysis_data/yearly_disease_count.csv")
-
-
-### Model data ####
 t = np.arange(2019, 2025, 1)
-DISEASES = ["Influenza", "Coronavirus", "Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
-SMALLER = ["Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
-LARGER = {"Coronavirus", "Respiratory", "Total"}
+KEYS = ["Coronavirus", "Respiratory", "Total"]
+df = pl.read_csv("data/02-analysis_data/yearly_disease_count.csv")
+raw_data = df.to_dict()
+for key in KEYS:
+  plt.plot(t, raw_data[key], marker = 'o', label = key)
 
-print(dict1)
-COLUMNS = ["Coronavirus", "Other", "Unknown"]
-for key in COLUMNS:
-  for i in range(6):
-    dict1[key][i] = 100*dict1[key][i]/dict1["Total Agents"][i]
+plt.title("Coronavirus, Respiratory & Total Outbreaks")
+plt.xlabel("Year")
+plt.ylabel("Outbreak Count")
+plt.grid()
 
-print(dict1)
-corona = np.array(dict1["Coronavirus"])
-other = np.array(dict1["Other"])
-unknown = np.array(dict1["Unknown"])
-
-ind = np.arange(6)    
-width = 0.35       
-
-p1 = plt.bar(ind, corona, width, color='#d62728', )
-p2 = plt.bar(ind, other, width,  bottom=corona)
-p3 = plt.bar(ind, unknown, width,  bottom=corona+other)
-
-plt.ylabel('Percentage of Causative Agents')
-plt.title('Composition of Respiratory Disease Outbreaks')
-plt.xticks(ind, ('2019', '2020', '2021', '2022', '2023', '2024'))
-plt.yticks(np.arange(0, 110, 10))
[…]

### 2025-05-13 paper/paper.qmd
got graphs & tables to work, still need reformatting

--- 2025-05-13 paper/paper.qmd
@@
 date-format: long
+
+execute:
+  echo: false
 
@@
 
-*insert table showing first few lines of raw data
+Below is an example of the raw data from 2019:
+
+```{python}
+#| label: fig-polar
+#| fig-cap: "A line plot on a polar axis"
+
+import polars as pl
+import matplotlib.pyplot as plt
+
+raw_data = pl.read_csv("../data/01-raw_data/ob_report_2019.csv")
+raw_data.head(5)
+
+```
 
@@
 
-*** insert table
+```{python}
+#| label: disease-count-preview
+#| fig-cap: "A line plot on a polar axis"
 
-Below are the summary statistics for the cleaned data:
+import polars as pl
+import matplotlib.pyplot as plt
 
-*** insert summary statistics
+raw_data = pl.read_csv("../data/02-analysis_data/disease_count.csv")
+raw_data = raw_data.head(5)
+raw_data
+```
+```{python}
+#| label: yearly-count-preview
+#| fig-cap: "A line plot on a polar axis"
+
+import polars as pl
+import matplotlib.pyplot as plt
+
+raw_data = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
+raw_data = raw_data.head(5)
+raw_data
+```
+
+Below are the summary statistics for the cleaned yearly count data:
+
+```{python}
+#| label: yearly-count-summary-statistics
+#| fig-cap: "A line plot on a polar axis"
+
+import polars as pl
+import matplotlib.pyplot as plt
+
+raw_data = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
+summary_data = raw_data.describe()
+summary_data
+```
 
@@
 Total & respiratory & Coronavirus:
-*** insert graph
+```{python}
+#| label: all-graph
+#| fig-cap: "A line plot on a polar axis"
+
+import polars as pl
+import matplotlib.pyplot as plt
+import numpy as np
+
+t = np.arange(2019, 2025, 1)
+KEYS = ["Coronavirus", "Respiratory", "Total", "Influenza", "Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
+df = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
+raw_data = df.to_dict()
+for key in KEYS:
+  plt.plot(t, raw_data[key], marker = 'o', label = key)
+
+plt.title("Coronavirus, Respiratory & Total Outbreaks")
+plt.xlabel("Year")
+plt.ylabel("Outbreak Count")
+plt.grid()
+
+plt.legend()
+plt.show()
+```
 
@@
 *** insert graph 
+```{python}
+#| label: composition-chart
+#| fig-cap: "A line plot on a polar axis"
+
+### Workspace setup ####
+import numpy as np
+import polars as pl
+import matplotlib.pyplot as plt
+
+#### Read data ####
+df = pl.read_csv("../data/02-analysis_data/disease_count.csv")
+disease_comp = df.to_dict()
+
+### Plot data ####
+t = np.arange(2019, 2025, 1)
+
+KEYS = ["Coronavirus", "Other", "Unknown"]
+
+for key in KEYS:
+  for i in range(6):
+    disease_comp[key][i] = 100*disease_comp[key][i]/disease_comp["Total Agents"][i]
+
+corona = np.array(disease_comp["Coronavirus"])
+other = np.array(disease_comp["Other"])
+unknown = np.array(disease_comp["Unknown"])
+
+ind = np.arange(6)    
+width = 0.35
+
+p1 = plt.bar(ind, corona, width, color='#d62728', )
+p2 = plt.bar(ind, other, width,  bottom=corona)
+p3 = plt.bar(ind, unknown, width,  bottom=corona+other)
+
+plt.ylabel('Percentage of Causative Agents')
+plt.title('Composition of Respiratory Disease Outbreaks')
+plt.xticks(ind, ('2019', '2020', '2021', '2022', '2023', '2024'))
+plt.yticks(np.arange(0, 110, 10))
+plt.legend((p1[0], p2[0], p3[0]), ('Coronavirus', 'Other', "Unknown"))
+
+plt.show()
+```
+
 Covid / Secondary Disease / Other & Unknown
@@
 
+```{python}
+#| label: no-corona-graph
+#| fig-cap: "A line plot on a polar axis"
+
+import polars as pl
+import matplotlib.pyplot as plt
+import numpy as np
+
+t = np.arange(2019, 2025, 1)
+KEYS = ["Influenza", "Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
+df = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
+raw_data = df.to_dict()
+for key in KEYS:
+  plt.plot(t, raw_data[key], marker = 'o', label = key)
+
+plt.title("Other Respiratory Outbreaks")
+plt.xlabel("Year")
+plt.ylabel("Outbreak Count")
+plt.grid()
+
+plt.legend()
+plt.show()
+```
+
 Parainfluenza, RSV, Metapneumovirus, and Rhinovirus:
+```{python}
+#| label: no-influenza-graph
+#| fig-cap: "A line plot on a polar axis"
+
+import polars as pl
+import matplotlib.pyplot as plt
+import numpy as np
+
+t = np.arange(2019, 2025, 1)
+KEYS = ["Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
+df = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
+raw_data = df.to_dict()
+for key in KEYS:
+  plt.plot(t, raw_data[key], marker = 'o', label = key)
+
+plt.title("Other Respiratory Outbreaks")
+plt.xlabel("Year")
+plt.ylabel("Outbreak Count")
+plt.grid()
+
+plt.legend()
+plt.show()
+```
+
 Each has specifics, see general trend of lower number & some w/ slight decrease reported in pre-pandemic & pandemic, with a subsequent increase post.

### 2025-05-13 scripts/05-exploratory_data_analysis.py

@@ -9,65 +9,12 @@
 
 import polars as pl
 import matplotlib.pyplot as plt
-import numpy as np
 
-t = np.arange(2019, 2025, 1)
-KEYS = ["Coronavirus", "Respiratory", "Total"]
-df = pl.read_csv("data/02-analysis_data/yearly_disease_count.csv")
-raw_data = df.to_dict()
-for key in KEYS:
-  plt.plot(t, raw_data[key], marker = 'o', label = key)
+raw_data = pl.read_csv("data/02-analysis_data/disease_count.csv")
+raw_data = raw_data.head(5)
+raw_data = raw_data.to_pandas()
+plt.table(raw_data)
 
-plt.title("Coronavirus, Respiratory & Total Outbreaks")
-plt.xlabel("Year")
-plt.ylabel("Outbreak Count")
-plt.grid()
-
-plt.legend()
 plt.show()
-
-
-#### Workspace setup ####
-# import numpy as np
-# import polars as pl
-# import matplotlib.pyplot as plt
-
-# #### Read data ####
-# composition_data = pl.read_csv("data/02-analysis_data/disease_count.csv")
-# dict1 = composition_data.to_dict(as_series = False)
-# analysis_data = pl.read_csv("data/02-analysis_data/yearly_disease_count.csv")
-
-
-# ### Model data ####
-# t = np.arange(2019, 2025, 1)
-# DISEASES = ["Influenza", "Coronavirus", "Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
-# SMALLER = ["Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
-# LARGER = {"Coronavirus", "Respiratory", "Total"}
-
-# print(dict1)
-# COLUMNS = ["Coronavirus", "Other", "Unknown"]
-# for key in COLUMNS:
-#   for i in range(6):
-#     dict1[key][i] = 100*dict1[key][i]/dict1["Total Agents"][i]
-
-# print(dict1)
-# corona = np.array(dict1["Coronavirus"])
-# other = np.array(dict1["Other"])
-# unknown = np.array(dict1["Unknown"])
-
-# ind = np.arange(6)    
-# width = 0.35       
-
-# p1 = plt.bar(ind, corona, width, color='#d62728', )
-# p2 = plt.bar(ind, other, width,  bottom=corona)
[…]

### 2025-05-13 paper/paper.qmd
Finalized figures

@@
 import matplotlib.pyplot as plt
+pl.Config.set_tbl_hide_column_data_types(True)  
+pl.Config(tbl_width_chars=1000)
 
 raw_data = pl.read_csv("../data/01-raw_data/ob_report_2019.csv")
+for key in raw_data.columns:
+  if key not in ["_id", "Type of Outbreak", "Causative Agent-1", "Causative Agent-2"]:
+    raw_data = raw_data.drop(key)
 raw_data.head(5)
@@
 import matplotlib.pyplot as plt
+pl.Config.set_tbl_hide_column_data_types(True)
 
@@
 raw_data = raw_data.head(5)
+raw_data = raw_data.rename({"Coronavirus": "Corona", "Syncytial Virus": "RSV",
+    "Metapneumovirus": "MPV",
+    "Influenza": "Flu", "Rhinovirus": "RV", "Parainfluenza": "PIV"})
 raw_data
@@
 import polars as pl
+import polars.selectors as sc
 import matplotlib.pyplot as plt
+pl.Config.set_tbl_hide_column_data_types(True)  
+pl.Config(tbl_width_chars=5)
 
@@
 summary_data = raw_data.describe()
-summary_data
+summary_data = summary_data.filter(pl.col("statistic") != "count")
+summary_data = summary_data.filter(pl.col("statistic") != "null_count")
+
+summary_data = summary_data.rename({"Coronavirus": "Corona", "Syncytial Virus": "RSV",
+    "Metapneumovirus": "MPV",
+    "Influenza": "Flu", "Rhinovirus": "RV", "Parainfluenza": "PIV"})
+
+summary_data = summary_data.with_columns(sc.by_dtype(pl.Float64).round(2))
+summary_data.drop("Year")
+
 ```
@@
 t = np.arange(2019, 2025, 1)
-KEYS = ["Coronavirus", "Respiratory", "Total", "Influenza", "Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
+KEYS = ["Total", "Respiratory", "Coronavirus", "Influenza", "Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
 df = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
@@
 
-plt.title("Coronavirus, Respiratory & Total Outbreaks")
+plt.title("Outbreak Counts")
 plt.xlabel("Year")
@@
 
-plt.legend()
+ax = plt.subplot(111)
+
+# Shrink current axis by 20%
+box = ax.get_position()
+ax.set_position([box.x0, box.y0, box.width * 0.8, box.height])
+
+# Put a legend to the right of the current axis
+ax.legend(loc='center left', bbox_to_anchor=(1, 0.5))
+
 plt.show()
@@
 
-plt.title("Other Respiratory Outbreaks")
+plt.title("Non-Coronavirus Respiratory Outbreaks")
 plt.xlabel("Year")
@@
 
-plt.legend()
-plt.show()
-```
+ax = plt.subplot(111)
 
-Parainfluenza, RSV, Metapneumovirus, and Rhinovirus:
-```{python}
-#| label: no-influenza-graph
-#| fig-cap: "A line plot on a polar axis"
+# Shrink current axis's height by 10% on the bottom
+box = ax.get_position()
+ax.set_position([box.x0, box.y0 + box.height * 0.1,
+                 box.width, box.height * 0.9])
 
-import polars as pl
-import matplotlib.pyplot as plt
-import numpy as np
+# Put a legend below current axis
+ax.legend(loc='upper center', bbox_to_anchor=(0.5, -0.05),
+          fancybox=True, shadow=True, ncol=5)
 
-t = np.arange(2019, 2025, 1)
-KEYS = ["Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
-df = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
-raw_data = df.to_dict()
-for key in KEYS:
-  plt.plot(t, raw_data[key], marker = 'o', label = key)
-
-plt.title("Other Respiratory Outbreaks")
-plt.xlabel("Year")
-plt.ylabel("Outbreak Count")
-plt.grid()
-
-plt.legend()
 plt.show()

### 2025-05-13 scripts/03-clean_data.py

@@ -19,6 +19,7 @@ DISEASE_COLUMNS = ["Coronavirus", "Influenza", "Syncytial Virus",
 
 #initializing dictionaries
 disease_count = dict()
+disease_count["Year"] = YEARS
 disease_count["Coronavirus"] = [0, ]*6
 disease_count["Other"] = [0, ]*6
 disease_count["Unknown"] = [0, ]*6

### 2025-05-13 paper/paper.qmd
rough draft of results

@@
 
-format: pdf
+format: 
+  pdf: 
+    fig-pos: "H"
 number-sections: true
@@
 
-Total & respiratory & Coronavirus:
+## Overall Cases:
 ```{python}
@@
 
-We see that the total number of cases fluctuates with the number of respiratory cases, with number of outbreaks increasing since 2019 to 2022, and beginning to decrease afterwards
+We see that the total number of cases fluctuates with the number of respiratory cases, with number of outbreaks increasing since 2019 to 2022, and beginning to decrease afterwards.
 
-The number of cases also fluctuates with the number of respiratory cases and total cases, following the same trend described above.
+The number of coronavirus cases also fluctuates with the number of respiratory cases and total cases, following the same trend described above.
 
-Composition of Respiratory Illness:
-*** insert graph 
+## Makeup of Respiratory Outbreaks
 ```{python}
 #| label: composition-chart
-#| fig-cap: "A line plot on a polar axis"
+#| fig-cap: "Makeup of Respiratory Outbreaks from 2019-2024 inclusive"
 
@@
 
-Covid / Secondary Disease / Other & Unknown
-
-Influenza:
-Number of cases fall from 2019- 2021, with 0 cases in 2021, and increase thereafter.
-
+It becomes evident that since the end of the pandemic, the proportion of Coronavirus outbreaks compared to non-coronavirus outbreaks has been steadily decreasing.
+## Non-Covid Respiratory Outbreaks
 ```{python}
@@
 
-Each has specifics, see general trend of lower number & some w/ slight decrease reported in pre-pandemic & pandemic, with a subsequent increase post.
+We see the number of Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus outbreaks all generally fell to lower levels during the pandemic (2020-2021). 
+Post pandemic (2022-2024), all, with the exception of Influenza, have resurged to greater than pre-pandemic (2019) levels.
+Influenza outbreaks have also increased since the pandemic, but remain at a lower level comparative to 2019.
 

### 2025-05-13 paper/paper.qmd
very rough draft of discussion

@@
 https://www.toronto.ca/wp-content/uploads/2023/06/9942-Outbreak-Preparedness-ToolkitRespiratoryTableTopFacilitators.pdf
+
+Healthcare facilities include ....
 
@@
 It becomes evident that since the end of the pandemic, the proportion of Coronavirus outbreaks compared to non-coronavirus outbreaks has been steadily decreasing.
+
 ## Non-Covid Respiratory Outbreaks
@@
 
-## Rise in Non-Covid Respiratory Outbreaks {#sec-first-point}
+It becomes apparent that given coronaviruses continue to comprise the greatest proportion of respiratory disease outbreaks in healthcare facilities in Toronto, and coronavirus outbreak trends match those of total respiratory outbreaks from, and total outbreaks from  2019-2024 that coronavirus outbreaks remain the biggest driver of disease outbreaks in Toronto.
+However, since the pandemic we have seen a decrease in both coronavirus cases, and percentage composition of all agents, indicating that its impact, while still pertinent, is decreasing.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+Simultaneously, other respiratory infections including Influenza, Parainfluenza, RSV, Metapneumovirus, and Rhinovirus have been experiencing a significant increase in outbreak cases.
+Though respiratory infections such as Influenza are well known to the public, those such as RSV are known to a much lesser extent with only 40% of adults having heard of RSV, and 30% who would definitely get the RSV vaccine.
+Compared to Influenza, this is a much lesser.
+(https://www.canada.ca/en/public-health/services/immunization-vaccines/vaccination-coverage/seasonal-influenza-survey-results-2023-2024.html).
+It is important to note that while these respiratory infections are less deadly than covid (https://www.nature.com/articles/s41598-024-55378-x ), many of these healthcare facilities are long term healthcare facilities, whose populations are more susceptible to these viruses.(https://www.cdc.gov/respiratory-viruses/risk-factors/older-adults.html#:~:text=As%20people%20get%20older%2C%20their,increasing%20sharply%20with%20advancing%20age.
+)
 
-## Decrease in Covid Percentage of Respiratory Outbreaks
-
-Please don't use these as sub-heading labels - change them to be what your point actually is.
+Altogether, this suggests a slightly greater emphasis should be placed on non-covid respiratory diseases, specifically RSV given it is the only one of the above which has a vaccine available which is not widely known, and outbreak numbers have risen in recent years post-pandemic.
+By campaigning for increased RSV vaccinations, when outbreaks do occur in healthcare facilities, they will be less deadly.
 
@@
 
-Weaknesses and next steps should also be included.
+The main limitation of this study is the scale.
+All data is at the scale of healthcare facilities, thus must remain in the context of such.
+Extension of these results to the general Toronto population may be misrepresentative, thus to confirm the same trends hold in the general public further research should be conducted with data coming from the general population.
 

### 2025-05-14 paper/paper.qmd
added references

@@
 
-On January 25, 2020, the first case of COVID-19 was reported in Canada, in a healthcare facility in Toronto (Cheese 2025).
+On January 25, 2020, the first case of COVID-19 was reported in Canada, in a healthcare facility in Toronto [@covidOverview].
 Around two months later on March 17, Doug Ford declared a state of emergency for Ontario, and the closure of facilities such as schools.
@@
 Schools went back in person, etc, etc, etc.
-(Compartin 2022).
+[@comartinIlunga2022].
 As the restrictions lift, it becomes important to characterize the effect they hold upon respiratory diseases both as a whole, and specific diseases.
 
-Previously, in 2021, research was conducted to examine the effects of covid protocol on other respiratory diseases, finding evidence for a decrease in disease during covid protocol (Groves et al 2021).
+Previously, in 2021, research was conducted to examine the effects of covid protocol on other respiratory diseases, finding evidence for a decrease in disease during covid protocol[@groves2021].
 However, there is a lack of literature examining the effects of the lifting of protocol on these same diseases, and examining the general trends.
@@
 Additionally, the percentage of respiratory outbreaks with covid as an agent has been decreasing since the pandemic.
-So, while an emphasis on covid specific protocols is important, preventative measures for other respiratory diseases, such as RSV vaccines, that were overshadowed during the pandemic should be further encouraged (Public Health Agency of Canada 2024).
+So, while an emphasis on covid specific protocols is important, preventative measures for other respiratory diseases, such as RSV vaccines, that were overshadowed during the pandemic should be further encouraged [@publicHealthAgency2024].
 
@@
 Outbreaks in Toronto Healthcare Institutions data is provided by Toronto Public Health., and is obtained from the OpenData Toronto Portal.
-The dataset is updated weekly on Thursdays with data as of the Wednesday at 2pm.(https://open.toronto.ca/dataset/outbreaks-in-toronto-healthcare-institutions/ ).
+The dataset is updated weekly on Thursdays with data as of the Wednesday at 2pm.[@outbreakData].
 
@@
 By the Ministry of Health, a respiratory outbreak is constituted by “Two cases of acute respiratory infections (ARI) within 48 hours with any common epidemiological link (e.g., unit, floor), at least one of which must be laboratory-confirmed; OR Three cases of ARI (laboratory confirmation not necessary) occurring within 48 hours with any common epidemiological link (e.g., unit, floor).
-https://www.toronto.ca/wp-content/uploads/2023/06/9942-Outbreak-Preparedness-ToolkitRespiratoryTableTopFacilitators.pdf
+[@torontoPublicHealth2023].
 
@@
 ## Analysis Data
+To simulate, test, download, and clean the Outbreaks in Toronto Healthcare Institutions (2019-2024), the statistical programming language Python was used [@python].
+Specific libraries that assisted the analysis include `numpy` [@numpy], `matplotlib` [@matplotlib], `polars` [@polars].
 
@@
 Compared to Influenza, this is a much lesser.
-(https://www.canada.ca/en/public-health/services/immunization-vaccines/vaccination-coverage/seasonal-influenza-survey-results-2023-2024.html).
-It is important to note that while these respiratory infections are less deadly than covid (https://www.nature.com/articles/s41598-024-55378-x ), many of these healthcare facilities are long term healthcare facilities, whose populations are more susceptible to these viruses.(https://www.cdc.gov/respiratory-viruses/risk-factors/older-adults.html#:~:text=As%20people%20get%20older%2C%20their,increasing%20sharply%20with%20advancing%20age.
-)
+[@torontoPublicHealth2023].
+It is important to note that while these respiratory infections are less deadly than covid [@grangier2024], many of these healthcare facilities are long term healthcare facilities, whose populations are more susceptible to these viruses [@cdc2024].
 
@@
 
-# References
+# References {-}

### 2025-05-14 scripts/02-download_data.py
fixed download & clean data scripts

@@ -10,7 +10,8 @@
 
 #### Workspace setup ####
 import polars as pl
-import io
+from io import StringIO
+import csv
 
 #### Download data ####
 import requests
@@ -28,18 +29,16 @@ params = { "id": "outbreaks-in-toronto-healthcare-institutions"}
 package = requests.get(url, params = params).json()
 
 # To get resource data:
-i = 2016
+i = 2025
 for idx, resource in enumerate(package["result"]["resources"]):
 
        # for datastore_active resources:
        if resource["datastore_active"]:
+            # To get all records in CSV format:
+            url = base_url + "/datastore/dump/" + resource["id"]
+            resource_dump_data = requests.get(url)
 
-           # To get all records in CSV format:
-           url = base_url + "/datastore/dump/" + resource["id"]
-           resource_dump_data = requests.get(url)
-           print("doing something for year " + str(i))
-
-           # Writing raw data to CSV
-           analysis_data = pl.DataFrame(io.StringIO(resource_dump_data.text))
-           analysis_data.write_csv("data/01-raw_data/" + str(i) + "_raw.csv", separator = ",")
-           i+=1
+            # Writing raw data to CSV
+            analysis_data = pl.read_csv(StringIO(resource_dump_data.text))
+            analysis_data.write_csv("data/01-raw_data/ob_report_" + str(i) + ".csv")
+            i-=1

### 2025-05-14 scripts/03-clean_data.py

@@ -68,7 +68,7 @@ for i in range(2019, 2025):
         else:
             disease_count["Other"][index] += 1
         
-        if disease_two != None and disease_two != "None":
+        if disease_two != None and disease_two != "" and disease_two != "None":
             disease_count["Total Agents"][index] += 1
             if "covid" in disease_two.lower() or "corona" in disease_two.lower():
                 disease_count["Coronavirus"][index] += 1

### 2025-05-14 paper/paper.qmd
fixed download & clean data scripts
finalized abstract & title

@@
 ---
-title: "Respiratory Disease Outbreaks in Toronto Pre, During, and Post Covid-19 Pandemic "
-subtitle: How Coronavirus and Other Respiratory Diseases have responded to the covid-19 pandemic and Toronto pandemic policies
+title: "Respiratory Outbreaks Trends in Toronto Healthcare Institutions Across the COVID-19 Pandemic "
+subtitle: "The Response of Coronavirus and Other Respiratory Diseases to COVID-19 Pandemic Policies"
 author: 
   - Evelyn Hughes
-thanks: "Code and data are available at: [https://github.com/EveHughes/Donaldson-Paper)."
+thanks: "Code and data are available at: [https://github.com/EveHughes/Donaldson-Paper]."
 date: 05/12/2025
@@
 
-abstract: "Through the analysis of Outbreaks in Toronto Healthcare Institutions, this paper seeks to identify the trends in respiratory infections in Toronto during the implementation and later lifting of covid-19 preventatory policies, including general trends in number of outbreaks, and makeup of respiratory infections.
-Through data visualization, this analysis finds that while reported coronavirus outbreaks have lowered since pandemic times, “secondary” respiratory diseases such as Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus that were at lower levels during the pandemic have begun to rise again.
-At the same time, covid becomes a lesser percentage of overall respiratory diseases.
-Given different respiratory diseases require different treatment, these findings are relevant for informing healthcare facilities to place a lesser emphasis on covid-19, etc."
+abstract: "Through the analysis of Outbreaks in Toronto Healthcare Institutions (2019-2025), this paper seeks to identify the trends in respiratory infections in Toronto healthcare institutions during the implementation and later lifting of COVID-19 preventative policies.
+Through data visualization this analysis finds that while reported Coronavirus outbreaks have decreased since the pandemic, non-Coronavirus respiratory outbreaks such as Influenza, Parainfluenza (PIV), Respiratory Syncytial Virus (RSV), Metapneumovirus (MPV), and Rhinovirus (RV) outbreaks have risen from the lower levels they were at during the pandemic.
+Concurrently, Coronaviruses have become a lesser percentage of respiratory causative agents in outbreaks.
+Given different respiratory diseases require different treatment, testing, and prevention, these findings are relevant for informing healthcare facilities of steps to take to mitigate and manage outbreaks."
 
@@
 ```{python}
-#| label: fig-polar
-#| fig-cap: "A line plot on a polar axis"
+#| label: table
+#| fig-cap: "2019 Raw Data"
 
@@
 To simulate, test, download, and clean the Outbreaks in Toronto Healthcare Institutions (2019-2024), the statistical programming language Python was used [@python].
-Specific libraries that assisted the analysis include `numpy` [@numpy], `matplotlib` [@matplotlib], `polars` [@polars].
+Specific libraries that assisted the analysis include `numpy` [@numpy], `matplotlib` [@matplotlib], and `polars` [@polars].
 

### 2025-05-14 scripts/00-simulate_data.py

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Simulates a dataset of Australian electoral divisions, including the 
+# Purpose: Simulates a yearly_dataset of Australian electoral divisions, including the 
   # state and party that won each division.
 # Author: Rohan Alexander
 # Date: 26 September 2024
@@ -21,31 +21,52 @@ np.random.seed(853)
 YEARS = ["2019", "2020", "2021", "2022", "2023", "2024"]
 DISEASE_COLUMNS = ["Coronavirus", "Influenza", "Syncytial Virus", 
                    "Metapneumovirus", "Rhinovirus", "Parainfluenza", "Respiratory", "Total"]
+COMPOSITION_COLUMNS = ["Coronavirus", "Other", "Unknown"]
 
-# Generate the outbreak data using numpy and polars
+# Generate the yearly outbreak data using numpy and polars
 #creating dictionary w/ columns
-data = dict()
-data["Year"] = YEARS
+yearly_data = dict()
+yearly_data["Year"] = YEARS
 for column in DISEASE_COLUMNS:
-  data[column] = [0, ]*6
-data["Respiratory"] =[0, ]*6
-data["Total"] = [0, ]*6
+  yearly_data[column] = [0, ]*6
+yearly_data["Respiratory"] =[0, ]*6
+yearly_data["Total"] = [0, ]*6
 
-#generating random values for data
+#generating random values for yearly data
 for i in range(6):
   respiratory_count = 0
   for disease in DISEASE_COLUMNS:
     outbreak_count = int(100*np.random.rand())
-    data[disease][i] = outbreak_count
+    yearly_data[disease][i] = outbreak_count
     respiratory_count += outbreak_count
 
   respiratory_count += int(100*np.random.rand())
-  data["Respiratory"][i] = respiratory_count
+  yearly_data["Respiratory"][i] = respiratory_count
   total_count = respiratory_count + int(100*np.random.rand())
-  data["Total"][i] = total_count
+  yearly_data["Total"][i] = total_count
 
-# Create a polars DataFrame
-analysis_data = pl.DataFrame(data)
 
-#### Save data ####
-analysis_data.write_csv("data/00-simulated_data/simulated_data.csv")
+# Generate disease count data using numpy and polars
+#creating dictionary w/ columns
+disease_count = dict()
+disease_count["Year"] = YEARS
+for column in COMPOSITION_COLUMNS:
+  disease_count[column] = [0, ]*6
[…]

### 2025-05-14 paper/paper.qmd
finalized introduction

@@
 
-On January 25, 2020, the first case of COVID-19 was reported in Canada, in a healthcare facility in Toronto [@covidOverview].
-Around two months later on March 17, Doug Ford declared a state of emergency for Ontario, and the closure of facilities such as schools.
-Additionally, mask mandates, social distancing guidelines, and vaccine cards were all put into place to mitigate the spread of the disease.
-Later around 2022, these restrictions began to lift.
-Schools went back in person, etc, etc, etc.
-[@comartinIlunga2022].
-As the restrictions lift, it becomes important to characterize the effect they hold upon respiratory diseases both as a whole, and specific diseases.
+On January 25, 2020, the first case of COVID-19 was reported in Canada in a healthcare facility in Toronto [@covidOverview].
+Around two months later on March 17, Doug Ford declared a state of emergency for Ontario and the closure of facilities such as schools.
+Mask mandates, social distancing guidelines, and vaccine cards were all put into place to mitigate the spread of the disease.
+Later around 2022, these restrictions slowly began to lift.
+On March 14, 2022 vaccination requirements were lifted, on March 21, 2022 social distancing regulations were lifted, and on April 27, 2022 all remaining COVID-19 regulatory policies and restrictions expired[@comartinIlunga2022].
+Given changes in outbreak count and type heavily impact healthcare facilities, it becomes important to characterize the effect of the lifting of COVID-19 regulations on respiratory diseases as a whole, and specific respiratory diseases.
 
-Previously, in 2021, research was conducted to examine the effects of covid protocol on other respiratory diseases, finding evidence for a decrease in disease during covid protocol[@groves2021].
-However, there is a lack of literature examining the effects of the lifting of protocol on these same diseases, and examining the general trends.
-This research aims to fill that gap, and identify how diseases have reacted to the lifting of covid protocol in relation to both their pre covid (2019), and during covid (2020-2022 ish) levels.
+Previously, in 2021, research was conducted to examine the effects of COVID-19 protocol on non-Coronavirus respiratory diseases, finding evidence for a decrease in disease during the restrictions [@groves2021].
+However, there is a lack of literature examining the effects of the lifting of protocol on these same respiratory diseases.
+This paper aims to fill that gap, and identify how diseases have reacted to the lifting of covid protocol in relation to both their pre covid (2019), and during covid (2020-2022 ish) levels.
 Doing such can inform policy that encourages disease-specific preventative measures, and promotion of vaccines to prevent increased outbreaks.
@@
 
-In the sections that follow are a description of the data, visualization of the data and discussion of the results.
+In what follows is a description of the data ([@sec-data]), visualization of the data ([@sec-results]), and an analysis of the visualization ([@sec-discussion]).
 
@@
 
-# Results
+# Results {#sec-results}
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 

### 2025-05-14 paper/paper.qmd
finalized data section

@@
 The dataset is updated weekly on Thursdays with data as of the Wednesday at 2pm.[@outbreakData].
+The dataset consists of data from years 2016-2025 inclusive, with each containing instances of outbreaks in toronto healthcare institutions during that year.
 
-The dataset consists of data from years 2016-2025 inclusive, with each containing instances of outbreaks in toronto healthcare institutions.
-
-“Under the Ontario Health Protection and Promotion Act (HPPA), healthcare institutions (hospitals, long-term care homes and retirement homes) are required to monitor staff and patients/residents for signs and symptoms of gastroenteric (e.g., nausea, vomiting, diarrhea, fever) and respiratory (e.g., cough, runny nose, sore throat, fever) infections.
-Healthcare institutions must also actively look for, detect and report suspected and/or confirmed outbreaks to their local public health unit”
-
-For every record there is an id number, institution name, institution address, outbreak setting (type of facility), type of outbreak, causative agent-1, causative agent-2, date outbreak began, date outbreak ended, and whether the outbreak is currently active.
-Below is a preview of this information:
-
-The type of outbreak is listed as Respiratory, Enteric, or Other.
-By the Ministry of Health, a respiratory outbreak is constituted by “Two cases of acute respiratory infections (ARI) within 48 hours with any common epidemiological link (e.g., unit, floor), at least one of which must be laboratory-confirmed; OR Three cases of ARI (laboratory confirmation not necessary) occurring within 48 hours with any common epidemiological link (e.g., unit, floor).
-[@torontoPublicHealth2023].
-
-Healthcare facilities include ....
-
-Causative Agent-1 is either a specific agent (ie: Coronavirus), or "Unable to identify".
-Causative Agent-2 is either a specific agent, or "None".
-
-Below is an example of the raw data from 2019:
+Each record includes an id number, institution name, institution address, outbreak setting (type of facility), type of outbreak, causative agent-1, causative agent-2, date outbreak began, date outbreak ended, and whether the outbreak is currently active.
+@tbl-raw_preview contains a preview of this information.
 
 ```{python}
-#| label: table
-#| fig-cap: "2019 Raw Data"
+#| label: tbl-raw_preview
+#| tbl-cap: "2019 Raw Data Preview"
 
@@
 
-For this analysis, the variables of interest are the yearly number of outbreaks for Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus.
-Over 2019-2024.
+Healthcare facilities are defined as any institution which provides a healthcare service, including hospitals and long term care homes.
+
+According to the Ontario Ministry of Health, a respiratory outbreak is defined as “Two cases of acute respiratory infections (ARI) within 48 hours with any common epidemiological link (e.g., unit, floor), at least one of which must be laboratory-confirmed; OR Three cases of ARI (laboratory confirmation not necessary) occurring within 48 hours with any common epidemiological link (e.g., unit, floor).
+[@torontoPublicHealth2023].
+
+The type of outbreak is categorized as Respiratory, Enteric, or Other. 
+Respiratory outbreaks refer to outbreaks that affect the respiratory system (eg. nose, lungs)
+Enteric outbreaks refer to outbreaks that affect the gastrointestinal tract.
+
+Causative Agent-1 is either a specific agent (ie: Coronavirus), or marked as "Unable to identify".
+Causative Agent-2 is either a specific agent, or marked as "".
+
+For this analysis, the variables of interest are the yearly number of outbreaks for Coronavirus/COVID-19, Influenza (Flu), Parainfluenza (PIV), Respiratory Syncytial Virus (RSV), Metapneumovirus (MPV), and Rhinovirus (RV) over the years of 2019-2024.
 
@@
 
-For each year, the number of cases were aggregated, and were aggregated within each respective category.
-The diseases of interest were: Coronavirus Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus.
-Instances of coronavirus and covid-19 were grouped together under the title of “coronavirus”, given covid is caused by a coronavirus.
-Influenza A & B were additionally grouped together under the title “Influenza”.
-
-Below is a sample of the cleaned data:
+For each year, the number of cases for each causative agent were aggregated.
+The causative agents of interest were: Coronavirus, Influenza, Parainfluenza (PIV), Respiratory Syncytial Virus (RSV), Metapneumovirus (MPV), and Rhinovirus (RV).
+Instances of Coronavirus and COVID-19 were grouped together under the category “Coronavirus”. Influenza A and Influenza B were additionally grouped together under the category “Influenza”. 
+@tbl-yearly_count_preview shows the data, and @tbl-yearly_count_summary shows the summary statistics.
 
 ```{python}
-#| label: disease-count-preview
-#| fig-cap: "A line plot on a polar axis"
-
-import polars as pl
-import matplotlib.pyplot as plt
-pl.Config.set_tbl_hide_column_data_types(True)
-
-raw_data = pl.read_csv("../data/02-analysis_data/disease_count.csv")
-raw_data = raw_data.head(5)
-raw_data
-```
-```{python}
-#| label: yearly-count-preview
-#| fig-cap: "A line plot on a polar axis"
+#| label: tbl-yearly_count_preview
+#| tbl-cap: "Yearly Outbreak Count"
 
@@
 
-Below are the summary statistics for the cleaned yearly count data:
-
 ```{python}
-#| label: yearly-count-summary-statistics
-#| fig-cap: "A line plot on a polar axis"
+#| label: tbl-yearly_count_summary
+#| tbl-cap: "Yearly Outbreak Count Summary Statistics"
 
@@
 
+Additionally, for each year the total number of non-null respiratory causative agents was aggregated, as well as the number of Coronavirus causative agents, non-Coronavirus respiratory causative agents, and unknown respiratory causative agents.
+@tbl-disease_count_preview shows the cleaned data, and @tbl-disease_count_summary shows the summary statistics.
+
+```{python}
+#| label: tbl-disease_count_preview
+#| tbl-cap: "Count of Respiratory Disease by Type"
+
+import polars as pl
+import matplotlib.pyplot as plt
+pl.Config.set_tbl_hide_column_data_types(True)
+
+raw_data = pl.read_csv("../data/02-analysis_data/disease_count.csv")
+raw_data = raw_data.head(6)
+raw_data
+```
+
+```{python}
+#| label: tbl-disease_count_summary
+#| tbl-cap: "Respiratory Disease Composition Summary Statistics"
+import polars as pl
+import polars.selectors as sc
+pl.Config.set_tbl_hide_column_data_types(True)
+
+raw_data = pl.read_csv("../data/02-analysis_data/disease_count.csv")
+summary_data = raw_data.describe()
+summary_data = summary_data.filter(pl.col("statistic") != "count")
+summary_data = summary_data.filter(pl.col("statistic") != "null_count")
+
+summary_data = summary_data.with_columns(sc.by_dtype(pl.Float64).round(2))
+summary_data.drop("Year")
+```
+
 # Results {#sec-results}
@@
 #| label: all-graph
-#| fig-cap: "A line plot on a polar axis"
+#| tbl-cap: "A line plot on a polar axis"
 

### 2025-05-14 paper/paper.qmd
Finalized results

@@
 ```{python}
-#| label: all-graph
-#| tbl-cap: "A line plot on a polar axis"
+#| label: fig-all_plot
+#| fig-cap: "Yearly Outbreak Count"
 
@@
 
-We see that the total number of cases fluctuates with the number of respiratory cases, with number of outbreaks increasing since 2019 to 2022, and beginning to decrease afterwards.
-
-The number of coronavirus cases also fluctuates with the number of respiratory cases and total cases, following the same trend described above.
+@fig-all_plot showcases that the total number of cases fluctuates with the number of respiratory cases, with number of outbreaks increasing since 2019 to 2022, and beginning to decrease afterwards.
+Additionally, the number of coronavirus cases also fluctuates with the number of respiratory cases and total cases, following the same trend described above.
 
@@
 ```{python}
-#| label: composition-chart
+#| label: fig-composition_chart
 #| fig-cap: "Makeup of Respiratory Outbreaks from 2019-2024 inclusive"
@@
 
-It becomes evident that since the end of the pandemic, the proportion of Coronavirus outbreaks compared to non-coronavirus outbreaks has been steadily decreasing.
+From @fig-composition_chart, it becomes evident that since the end of the pandemic the proportion of Coronavirus outbreaks compared to Non-Coronavirus outbreaks has been steadily decreasing.
 
@@
 ```{python}
-#| label: no-corona-graph
-#| fig-cap: "A line plot on a polar axis"
+#| label: fig-no_corona_graph
+#| fig-cap: "Non-Coronavirus Outbreaks"
 
@@
 
-We see the number of Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus outbreaks all generally fell to lower levels during the pandemic (2020-2021). 
-Post pandemic (2022-2024), all, with the exception of Influenza, have resurged to greater than pre-pandemic (2019) levels.
-Influenza outbreaks have also increased since the pandemic, but remain at a lower level comparative to 2019.
+From @fig-no_corona_graph,  it can be observed that the number of Influenza, Parainfluenza, Respiratory Syncytial Virus, Metapneumovirus, and Rhinovirus outbreaks all generally fell to lower levels during the pandemic (2020-2021).
+Post-pandemic (2022-2024), all, with the exception of Influenza, have resurged to greater than pre-pandemic (2019) levels.
+Influenza outbreaks have also increased since the pandemic, but remain at a lower level comparative to their 2019 pre-pandemic level.
 

### 2025-05-14 paper/paper.qmd
Finalized discussion

@@
 
-It becomes apparent that given coronaviruses continue to comprise the greatest proportion of respiratory disease outbreaks in healthcare facilities in Toronto, and coronavirus outbreak trends match those of total respiratory outbreaks from, and total outbreaks from  2019-2024 that coronavirus outbreaks remain the biggest driver of disease outbreaks in Toronto.
-However, since the pandemic we have seen a decrease in both coronavirus cases, and percentage composition of all agents, indicating that its impact, while still pertinent, is decreasing.
+It is apparent that given Coronaviruses outbreaks remain the most prominent driver of outbreaks in Toronto healthcare facilities post-pandemic.
+They continue to comprise the greatest proportion of respiratory disease outbreaks in healthcare facilities in Toronto, and since pandemic times, trends in Coronavirus outbreak count align with those of total respiratory outbreaks and total outbreaks.
+However, since the pandemic the annual number of Coronavirus outbreaks and percentage Coronavirus outbreaks comprising total respiratory outbreak agents has declined, indicating that their impact is diminishing.
 
-Simultaneously, other respiratory infections including Influenza, Parainfluenza, RSV, Metapneumovirus, and Rhinovirus have been experiencing a significant increase in outbreak cases.
-Though respiratory infections such as Influenza are well known to the public, those such as RSV are known to a much lesser extent with only 40% of adults having heard of RSV, and 30% who would definitely get the RSV vaccine.
-Compared to Influenza, this is a much lesser.
-[@torontoPublicHealth2023].
-It is important to note that while these respiratory infections are less deadly than covid [@grangier2024], many of these healthcare facilities are long term healthcare facilities, whose populations are more susceptible to these viruses [@cdc2024].
+Simultaneously, Influenza, Parainfluenza, RSV, Metapneumovirus, and Rhinovirus have been experiencing an increase in outbreak cases post-pandemic.
+While Influenza is well known, the public is much less familiar with other respiratory outbreak agents, particularly RSV.
+Only 40% of adults report having heard of RSV, and only 30% would definitely get the RSV vaccine.
+Comparatively, 43% of adults got an Influenza vaccine in the 2023-2024 season [@torontoPublicHealth2023].
+It is important to note that while non-Coronavirus respiratory infections such as RSV are less deadly than Coronavirus [@grangier2024], many of the healthcare facilities experiencing outbreaks are long term healthcare facilities whose populations are more susceptible to these infections [@cdc2024].
 
-Altogether, this suggests a slightly greater emphasis should be placed on non-covid respiratory diseases, specifically RSV given it is the only one of the above which has a vaccine available which is not widely known, and outbreak numbers have risen in recent years post-pandemic.
-By campaigning for increased RSV vaccinations, when outbreaks do occur in healthcare facilities, they will be less deadly.
+The lack of knowledge surrounding non-Coronavirus and Influenza respiratory causative agents, their rising outbreak numbers, and susceptibility of healthcare institution population suggests a greater emphasis should be placed on them.
+RSV in particular should be given a greater emphasis, given it is the only one of the non-Coronavirus and non-Influenza respiratory agents with a vaccine available.
+By increasing awareness of RSV and encouraging RSV vaccinations, RSV outbreaks will be less likely to occur in healthcare facilities, and when they do occur, they will be less severe.
 
@@
 
-The main limitation of this study is the scale.
-All data is at the scale of healthcare facilities, thus must remain in the context of such.
-Extension of these results to the general Toronto population may be misrepresentative, thus to confirm the same trends hold in the general public further research should be conducted with data coming from the general population.
+The main limitation of this study is the limitation of focus to healthcare facilities.
+All data is from healthcare facilities, thus must remain in the context of such.
+The extension of these results to the general Toronto population may be misrepresentative, thus to confirm the same trends hold in the general public, further research should be conducted with data coming from the general Toronto public.
 

### 2025-05-14 paper/paper.qmd
Finalized paper

@@
 Concurrently, Coronaviruses have become a lesser percentage of respiratory causative agents in outbreaks.
-Given different respiratory diseases require different treatment, testing, and prevention, these findings are relevant for informing healthcare facilities of steps to take to mitigate and manage outbreaks."
+Given that different respiratory diseases require different treatment, testing, and prevention, these findings are relevant for informing healthcare facilities of steps to take to mitigate and manage outbreaks."
 
@@
 On January 25, 2020, the first case of COVID-19 was reported in Canada in a healthcare facility in Toronto [@covidOverview].
-Around two months later on March 17, Doug Ford declared a state of emergency for Ontario and the closure of facilities such as schools.
-Mask mandates, social distancing guidelines, and vaccine cards were all put into place to mitigate the spread of the disease.
-Later around 2022, these restrictions slowly began to lift.
-On March 14, 2022 vaccination requirements were lifted, on March 21, 2022 social distancing regulations were lifted, and on April 27, 2022 all remaining COVID-19 regulatory policies and restrictions expired[@comartinIlunga2022].
-Given changes in outbreak count and type heavily impact healthcare facilities, it becomes important to characterize the effect of the lifting of COVID-19 regulations on respiratory diseases as a whole, and specific respiratory diseases.
+Around two months later, on March 17, Doug Ford declared a state of emergency for Ontario and the closure of facilities such as schools.
+Mask mandates, social distancing guidelines, and vaccine cards were all implemented to mitigate the spread of the disease.
+Later, around 2022, these restrictions slowly began to lift.
+On March 14, 2022, vaccination requirements were lifted, on March 21, 2022, social distancing regulations were lifted, and on April 27, 2022, all remaining COVID-19 regulatory policies and restrictions expired[@comartinIlunga2022].
+Given that changes in outbreak count and type heavily impact healthcare facilities, it becomes important to characterize the effect of the lifting of COVID-19 regulations on respiratory diseases as a whole, as well as specific respiratory diseases.
 
 Previously, in 2021, research was conducted to examine the effects of COVID-19 protocol on non-Coronavirus respiratory diseases, finding evidence for a decrease in disease during the restrictions [@groves2021].
-However, there is a lack of literature examining the effects of the lifting of protocol on these same respiratory diseases.
-This paper aims to fill that gap, and identify how diseases have reacted to the lifting of covid protocol in relation to both their pre covid (2019), and during covid (2020-2022 ish) levels.
-Doing such can inform policy that encourages disease-specific preventative measures, and promotion of vaccines to prevent increased outbreaks.
+However, there is a lack of literature examining the effects of lifting the protocol on these same respiratory diseases.
+This paper aims to improve upon that deficit and identify how diseases have reacted to the lifting of COVID-19 protocol in relation to both their pre-pandemic (2019) and pandemic (2020-2022) levels.
+Doing such can inform policy that encourages disease-specific preventative measures and promotion of vaccines to prevent increased outbreaks.
 
@@
 Then, it looks at coronavirus, Influenza (A & B), Parainfluenza, Respiratory Syncytial Virus (RSV), Metapneumovirus, and Rhinovirus.
-Finally, it examines the trends in percentage of respiratory outbreaks with covid as an agent.
-Through the data visualization, it was found that while coronavirus outbreaks have been falling since their covid levels, other respiratory diseases have been seeing a subsequent increase in cases post pandemic.
-Additionally, the percentage of respiratory outbreaks with covid as an agent has been decreasing since the pandemic.
-So, while an emphasis on covid specific protocols is important, preventative measures for other respiratory diseases, such as RSV vaccines, that were overshadowed during the pandemic should be further encouraged [@publicHealthAgency2024].
+Finally, it examines the trends in the percentage of respiratory outbreaks with COVID-19 or Coronavirus as an agent.
 
-In what follows is a description of the data ([@sec-data]), visualization of the data ([@sec-results]), and an analysis of the visualization ([@sec-discussion]).
+Through the data visualization, it was found that while Coronavirus outbreaks have been falling since their pandemic levels, other respiratory diseases have been seeing a subsequent increase in cases post-pandemic.
+Additionally, the percentage of respiratory outbreaks with COVID-19 or Coronavirus as an agent has decreased since the pandemic.
+So, while an emphasis on COVID-specific protocols is important, preventative measures for other respiratory diseases, such as RSV vaccines, that were overshadowed during the pandemic should be further encouraged [@publicHealthAgency2024].
+
+In what follows is a description of the data ([@sec-data]), a visualization of the data ([@sec-results]), and an analysis of the visualization ([@sec-discussion]).
 
@@
 
-Outbreaks in Toronto Healthcare Institutions data is provided by Toronto Public Health., and is obtained from the OpenData Toronto Portal.
-The dataset is updated weekly on Thursdays with data as of the Wednesday at 2pm.[@outbreakData].
-The dataset consists of data from years 2016-2025 inclusive, with each containing instances of outbreaks in toronto healthcare institutions during that year.
+Outbreaks in Toronto Healthcare Institutions data is provided by Toronto Public Health, and is obtained from the OpenData Toronto Portal.
+The dataset is updated weekly on Thursdays [@outbreakData].
+The dataset consists of data from years 2016-2025 inclusive, with each containing instances of outbreaks in Toronto healthcare institutions during that year.
 
-Each record includes an id number, institution name, institution address, outbreak setting (type of facility), type of outbreak, causative agent-1, causative agent-2, date outbreak began, date outbreak ended, and whether the outbreak is currently active.
+Each record includes an ID number, institution name, institution address, outbreak setting (type of facility), type of outbreak, causative agent-1, causative agent-2, date outbreak began, date outbreak ended, and whether the outbreak is currently active.
 @tbl-raw_preview contains a preview of this information.
@@
 
+#set up
 import polars as pl
@@
 
+#reading in of data & removing unneeded info
 raw_data = pl.read_csv("../data/01-raw_data/ob_report_2019.csv")
@@
     raw_data = raw_data.drop(key)
+
+#creating table
 raw_data.head(5)
@@
 
-Healthcare facilities are defined as any institution which provides a healthcare service, including hospitals and long term care homes.
+Healthcare facilities are defined as any institution that provides a healthcare service, including hospitals and long-term care homes.
 
@@
 The type of outbreak is categorized as Respiratory, Enteric, or Other. 
-Respiratory outbreaks refer to outbreaks that affect the respiratory system (eg. nose, lungs)
+Respiratory outbreaks refer to outbreaks that affect the respiratory system.
 Enteric outbreaks refer to outbreaks that affect the gastrointestinal tract.
@@
 Causative Agent-1 is either a specific agent (ie: Coronavirus), or marked as "Unable to identify".
-Causative Agent-2 is either a specific agent, or marked as "".
+Causative Agent-2 is either a specific agent or marked as "".
 
-For this analysis, the variables of interest are the yearly number of outbreaks for Coronavirus/COVID-19, Influenza (Flu), Parainfluenza (PIV), Respiratory Syncytial Virus (RSV), Metapneumovirus (MPV), and Rhinovirus (RV) over the years of 2019-2024.
+For this analysis, the variables of interest are the yearly number of outbreaks for Coronavirus/COVID-19, Influenza (Flu), Parainfluenza (PIV), Respiratory Syncytial Virus (RSV), Metapneumovirus (MPV), and Rhinovirus (RV) over the years 2019-2024.
 
 ## Analysis Data
+### Yearly Outbreak Count {#sec-yearly_count}
 To simulate, test, download, and clean the Outbreaks in Toronto Healthcare Institutions (2019-2024), the statistical programming language Python was used [@python].
@@
 
-For each year, the number of cases for each causative agent were aggregated.
+For each year, the number of cases for each causative agent was aggregated.
 The causative agents of interest were: Coronavirus, Influenza, Parainfluenza (PIV), Respiratory Syncytial Virus (RSV), Metapneumovirus (MPV), and Rhinovirus (RV).
@@
 
+#reading in & reformatting
 raw_data = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
@@
     "Influenza": "Flu", "Rhinovirus": "RV", "Parainfluenza": "PIV"})
+
+#creating table
 raw_data
@@
 
+#rounding data
 summary_data = summary_data.with_columns(sc.by_dtype(pl.Float64).round(2))
+
+#creating table
 summary_data.drop("Year")
@@
 
-Additionally, for each year the total number of non-null respiratory causative agents was aggregated, as well as the number of Coronavirus causative agents, non-Coronavirus respiratory causative agents, and unknown respiratory causative agents.
+It is important to note that in @tbl-yearly_count_preview, non-Coronavirus respiratory causative agents in 2021 had extremely low outbreak counts.
+While this could be representative of genuinely surpressed levels of other respiratory outbreaks due to increased COVID-19 cases, it could be indicative of a reporting issue.
+Further research should be conducted to clarify, though such research is beyond the scope of this paper.
+
+### Composition Respiratory Disease Causative Agents
+Additionally, for each year, the total number of non-null respiratory causative agents was aggregated, as well as the number of Coronavirus causative agents, non-Coronavirus respiratory causative agents, and unknown respiratory causative agents.
 @tbl-disease_count_preview shows the cleaned data, and @tbl-disease_count_summary shows the summary statistics.
@@
 #| label: tbl-disease_count_preview
-#| tbl-cap: "Count of Respiratory Disease by Type"
+#| tbl-cap: "Count of Respiratory Disease by Coronavirus, Non-Coronavirus, or Other"
 
+#set up
 import polars as pl
@@
 
+#reading in data & creating table
 raw_data = pl.read_csv("../data/02-analysis_data/disease_count.csv")
@@
 #| tbl-cap: "Respiratory Disease Composition Summary Statistics"
+
+#set up
 import polars as pl
@@
 
+#reading in data & reformatting
 raw_data = pl.read_csv("../data/02-analysis_data/disease_count.csv")
@@
 
+#rounding all data
 summary_data = summary_data.with_columns(sc.by_dtype(pl.Float64).round(2))
+
+#creating table
 summary_data.drop("Year")
@@
 
+#plotting data & adding labels
 t = np.arange(2019, 2025, 1)
@@
   plt.plot(t, raw_data[key], marker = 'o', label = key)
-
 plt.title("Outbreak Counts")
@@
 
+# formatting legend
 ax = plt.subplot(111)
-
-# Shrink current axis by 20%
 box = ax.get_position()
 ax.set_position([box.x0, box.y0, box.width * 0.8, box.height])
-
-# Put a legend to the right of the current axis
 ax.legend(loc='center left', bbox_to_anchor=(1, 0.5))
 
+#showing
 plt.show()
@@
 @fig-all_plot showcases that the total number of cases fluctuates with the number of respiratory cases, with number of outbreaks increasing since 2019 to 2022, and beginning to decrease afterwards.
-Additionally, the number of coronavirus cases also fluctuates with the number of respiratory cases and total cases, following the same trend described above.
+Additionally, the number of Coronavirus cases also fluctuates with the number of respiratory cases and total cases, following the same trend described above.
 
@@
 
-### Workspace setup ####
+#set up
 import numpy as np
@@
 
-#### Read data ####
+# read data
 df = pl.read_csv("../data/02-analysis_data/disease_count.csv")
@@
 
-### Plot data ####
+# Plot data
 t = np.arange(2019, 2025, 1)
@@
 
+# set up
 import polars as pl
@@
 
+#plot data
 t = np.arange(2019, 2025, 1)
@@
 
+# Set up legend
 ax = plt.subplot(111)
-
-# Shrink current axis's height by 10% on the bottom
 box = ax.get_position()
@@
                  box.width, box.height * 0.9])
-
-# Put a legend below current axis
 ax.legend(loc='upper center', bbox_to_anchor=(0.5, -0.05),
@@
 
-It is apparent that given Coronaviruses outbreaks remain the most prominent driver of outbreaks in Toronto healthcare facilities post-pandemic.
-They continue to comprise the greatest proportion of respiratory disease outbreaks in healthcare facilities in Toronto, and since pandemic times, trends in Coronavirus outbreak count align with those of total respiratory outbreaks and total outbreaks.
-However, since the pandemic the annual number of Coronavirus outbreaks and percentage Coronavirus outbreaks comprising total respiratory outbreak agents has declined, indicating that their impact is diminishing.
+It is apparent that Coronavirus outbreaks remain the most prominent driver of outbreaks in Toronto healthcare facilities post-pandemic.
+They continue to comprise the greatest proportion of respiratory disease outbreaks in healthcare facilities in Toronto, and since pandemic times, trends in the Coronavirus outbreak count align with those of total respiratory outbreaks and total outbreaks.
+However, since the pandemic, the annual number of Coronavirus outbreaks and percentage of Coronavirus outbreaks comprising total respiratory outbreak agents has declined, indicating that their impact is diminishing.
 
@@
 Comparatively, 43% of adults got an Influenza vaccine in the 2023-2024 season [@torontoPublicHealth2023].
-It is important to note that while non-Coronavirus respiratory infections such as RSV are less deadly than Coronavirus [@grangier2024], many of the healthcare facilities experiencing outbreaks are long term healthcare facilities whose populations are more susceptible to these infections [@cdc2024].
+It is important to note that while non-Coronavirus respiratory infections, such as RSV, are less deadly than Coronavirus [@grangier2024], many of the healthcare facilities experiencing outbreaks are long-term healthcare facilities whose populations are more susceptible to these infections [@cdc2024].
 
-The lack of knowledge surrounding non-Coronavirus and Influenza respiratory causative agents, their rising outbreak numbers, and susceptibility of healthcare institution population suggests a greater emphasis should be placed on them.
+The lack of knowledge surrounding non-Coronavirus and Influenza respiratory causative agents, their rising outbreak numbers, and susceptibility of the healthcare institution population suggests a greater emphasis should be placed on them.
 RSV in particular should be given a greater emphasis, given it is the only one of the non-Coronavirus and non-Influenza respiratory agents with a vaccine available.
@@
 
-The main limitation of this study is the limitation of focus to healthcare facilities.
-All data is from healthcare facilities, thus must remain in the context of such.
-The extension of these results to the general Toronto population may be misrepresentative, thus to confirm the same trends hold in the general public, further research should be conducted with data coming from the general Toronto public.
+The main limitation of this study is the restriction of focus to healthcare facilities.
+All data is from healthcare facilities, hence, findings must remain in the context of healthcare facilities.
+The extension of these results to the general Toronto population may be misrepresentative.
+Thus, to confirm the same trends hold in general, further research should be conducted with data coming from the general Toronto public.
+
+Additionally, as mentioned in [@sec-yearly_count], the extremely low reported outbreak counts of non-Coronavirus causative agents in 2021 point to potential issues with the validity of the data.
+Further research should be conducted to determine the cause of these low numbers and assess the validity of the dataset.
 

### 2025-05-14 scripts/05-exploratory_data_analysis.py
removed uneccessary files & filled out preamble

@@ -1,20 +1,0 @@
-#### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
-import polars as pl
-import matplotlib.pyplot as plt
-
-raw_data = pl.read_csv("data/02-analysis_data/disease_count.csv")
-raw_data = raw_data.head(5)
-raw_data = raw_data.to_pandas()
-plt.table(raw_data)
-
-plt.show()
-
-#### Save model ####

### 2025-05-14 scripts/00-simulate_data.py
updated preambles

@@ -1,9 +1,9 @@
 #### Preamble ####
 # Purpose: Simulates a yearly_dataset of Australian electoral divisions, including the 
   # state and party that won each division.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Author: Evelyn Hughes
+# Date: 13 May 2025
+# Contact: evelyn.hughes@utoronto.ca
 # License: MIT
 # Pre-requisites: 
   # - `polars` must be installed (pip install polars)

### 2025-05-14 scripts/01-test_simulated_data.py

@@ -1,12 +1,11 @@
 #### Preamble ####
-# Purpose: Tests the structure and validity of the simulated Australian 
-  #electoral divisions dataset.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Tests the structure and validity of the simulated data sets
+# Author: Evelyn Hughes
+# Date: 13 May 2025
+# Contact: evelyn.hughes@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: 
   # - The `tidyverse` package must be installed and loaded
-  # - 00-simulate_data.R must have been run
+  # - 00-simulate_data.py must have been run
 # Any other information needed? Make sure you are in the `starter_folder` rproj
 

### 2025-05-14 scripts/02-download_data.py

@@ -1,17 +1,17 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Downloads and saves the data from Outbreaks in Toronto Healthcare Institutions
+# Author: Evelyn Hughes
+# Date: 14 May 2025
+# Contact: evelyn.hughes@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: 
+  # - `polars` must be installed (pip install polars)
+  # - `numpy` must be installed (pip install numpy)
+  # - `io` must be installed (pip install io)
 
 #### Workspace setup ####
 import polars as pl
 from io import StringIO
-import csv
 
 #### Download data ####
 import requests

### 2025-05-14 scripts/03-clean_data.py

@@ -1,11 +1,12 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Cleans the raw data from Outbreaks in Toronto Healthcare Institutions
+# Author: Evelyn Hughes
+# Date: 13 May 2025
+# Contact: evelyn.hughes@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 
+  # -  02-download_data.py must have been run
+  # - `polars` must be installed (pip install polars)
 
 #### Workspace setup ####
 import polars as pl

### 2025-05-14 scripts/04-test_analysis_data.py

@@ -1,65 +1,0 @@
-#### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 26 September 2024 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
-
-#### Workspace setup ####
-import polars as pl
-disease_count = pl.read_csv("data/02-analysis_data/disease_count.csv")
-yearly_disease_count = pl.read_csv("data/02-analysis_data/yearly_disease_count.csv")
-
-
-#### Test disease_count data ####
-# Test that there are no missing values in the dataset
-assert disease_count.null_count().to_series().sum() == 0, "Dataset has missing values"
-
-#Test all columns are int type
-for type in disease_count.dtypes:
-  assert type == pl.Int64
-
-#Test counts add to total
-for row in disease_count.rows(named = True):
-  assert row["Coronavirus"] + row["Other"] + row["Unknown"] == row["Total Agents"], "sum does not equal total"
-
-#### Test yearly_disease_count data ####
-# Test that the dataset has 10 rows - one for each year between 2016-2025 inclusive
-assert yearly_disease_count.shape[0] == 6, "Dataset does not have 10 rows"
-
-# Test that the dataset has 6 columns
-assert yearly_disease_count.shape[1] == 9, "Dataset does not have 9 columns"
-
-# Test that the 'Coronavirus' column is int
-assert yearly_disease_count["Coronavirus"].dtype == pl.Int64, "Coronavirus column is not int type"
-
-# Test that the 'Influenza' column is int 
-assert yearly_disease_count["Influenza"].dtype == pl.Int64, "Influenza column is not int type"
-
-# Test that the 'Syncytial Virus' column is int 
-assert yearly_disease_count["Syncytial Virus"].dtype == pl.Int64, "Syncytial Virus column is not int type"
-
-# Test that the 'Metapneumovirus' column is int
-assert yearly_disease_count["Metapneumovirus"].dtype == pl.Int64, "Metapneumovirus column is not int type"
-
-# Test that the 'Rhinovirus' column is int
-assert yearly_disease_count["Rhinovirus"].dtype == pl.Int64, "Rhinovirus column is not int type"
-
-# Test that the 'Parainfluenza' column is int
[…]

### 2025-05-14 scripts/test_analysis_data.py

@@ -1,0 +1,77 @@
+#### Preamble ####
+# Purpose: Tests Cleaned Disease Count Data & Yearly Disease Count Data
+# Author: Evelyn Hughes
+# Date: 13 May 2025
+# Contact: evelyn.hughes@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: 
+  # - `polars` must be installed (pip install polars)
+  # -  02-download_data.py must have been run
+  # -  03-clean_data.py must have been run
+
+
+#### Workspace setup ####
+import polars as pl
+import unittest
+
+if __name__ == "__main__":
+  
+  unittest.main()
+
+
+#### Test disease_count data ####
+class TestAll(unittest.TestCase):
+  def setUp(cls):
+    cls.disease_count = pl.read_csv("data/02-analysis_data/disease_count.csv")
+    cls.yearly_disease_count = pl.read_csv("data/02-analysis_data/yearly_disease_count.csv")
+
+  def test_disease_count_data(self):
+    # Test that there are no missing values in the dataset
+    self.assertEqual(self.disease_count.null_count().to_series().sum(), 0)
+
+    #Test all columns are int type
+    for type in self.disease_count.dtypes:
+      self.assertEqual(type, pl.Int64)
+
+    #Test counts add to total
+    for row in self.disease_count.rows(named = True):
+      self.assertEqual(row["Coronavirus"] + row["Other"] + row["Unknown"], row["Total Agents"])
+
+#### Test yearly_disease_count data ####
+  def test_yearly_disease_count():
+    ## Test that the dataset has 10 rows - one for each year between 2016-2025 inclusive
+    assert yearly_disease_count.shape[0] == 6, "Dataset does not have 6 rows"
+
+    # Test that the dataset has 6 columns
+    assert yearly_disease_count.shape[1] == 9, "Dataset does not have 9 columns"
+
+    # Test that the 'Coronavirus' column is int
+    assert yearly_disease_count["Coronavirus"].dtype == pl.Int64, "Coronavirus column is not int type"
+
+    # Test that the 'Influenza' column is int 
+    assert yearly_disease_count["Influenza"].dtype == pl.Int64, "Influenza column is not int type"
+
+    # Test that the 'Syncytial Virus' column is int 
+    assert yearly_disease_count["Syncytial Virus"].dtype == pl.Int64, "Syncytial Virus column is not int type"
+
+    # Test that the 'Metapneumovirus' column is int
[…]

### 2025-05-14 scripts/00-simulate_data.py
Got tests running

@@ -52,7 +52,7 @@ disease_count = dict()
 disease_count["Year"] = YEARS
 for column in COMPOSITION_COLUMNS:
   disease_count[column] = [0, ]*6
-disease_count["Total"] = [0, ]*6
+disease_count["Total Agents"] = [0, ]*6
 
 #generating random values for disease count
 for i in range(6):
@@ -61,7 +61,7 @@ for i in range(6):
     outbreak_count = int(100*np.random.rand())
     disease_count[disease][i] = outbreak_count
     total_count += outbreak_count
-  disease_count["Total"][i] = total_count
+  disease_count["Total Agents"][i] = total_count
 
 # Create a polars yearly_dataFrame
 analysis_yearly_data = pl.DataFrame(yearly_data)

### 2025-05-14 scripts/01-test_simulated_data.py

@@ -1,11 +1,0 @@
-#### Preamble ####
-# Purpose: Tests the structure and validity of the simulated data sets
-# Author: Evelyn Hughes
-# Date: 13 May 2025
-# Contact: evelyn.hughes@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: 
-  # - The `tidyverse` package must be installed and loaded
-  # - 00-simulate_data.py must have been run
-# Any other information needed? Make sure you are in the `starter_folder` rproj
-

### 2025-05-14 scripts/test_simulated_data.py

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Tests Cleaned Disease Count Data & Yearly Disease Count Data
+# Purpose: Tests Simulated Data
 # Author: Evelyn Hughes
 # Date: 13 May 2025
 # Contact: evelyn.hughes@mail.utoronto.ca
@@ -7,18 +7,16 @@
 # Pre-requisites: 
   # - `polars` must be installed (pip install polars)
   # - 'unittest' must be installed
-  # -  02-download_data.py must have been run
-  # -  03-clean_data.py must have been run
-
+  # -  00-simulate_data.py must have been run
 
 #### Workspace setup ####
 import polars as pl
 import unittest
 
-class TestAll(unittest.TestCase):
+class TestSimulated(unittest.TestCase):
   def setUp(self):
-    self.disease_count = pl.read_csv("data/02-analysis_data/disease_count.csv")
-    self.yearly_disease_count = pl.read_csv("data/02-analysis_data/yearly_disease_count.csv")
+    self.disease_count = pl.read_csv("data/00-simulated_data/simulated_disease_count.csv")
+    self.yearly_disease_count = pl.read_csv("data/00-simulated_data/simulated_yearly_count.csv")
 
   def test_diseaseCount(self):
 

### 2025-05-14 paper/paper.qmd
Fixed typos

@@
 ---
-title: "Respiratory Outbreaks Trends in Toronto Healthcare Institutions Across the COVID-19 Pandemic "
+title: "Respiratory Outbreak Trends in Toronto Healthcare Institutions Across the COVID-19 Pandemic "
 subtitle: "The Response of Coronavirus and Other Respiratory Diseases to COVID-19 Pandemic Policies"
@@
 ### Yearly Outbreak Count {#sec-yearly_count}
-To simulate, test, download, and clean the Outbreaks in Toronto Healthcare Institutions (2019-2024), the statistical programming language Python was used [@python].
+To simulate, test, download, and clean the Outbreaks in Toronto Healthcare Institutions (2019-2024), the programming language Python was used [@python].
 Specific libraries that assisted the analysis include `numpy` [@numpy], `matplotlib` [@matplotlib], and `polars` [@polars].
@@
 t = np.arange(2019, 2025, 1)
-KEYS = ["Total", "Respiratory", "Coronavirus", "Influenza", "Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
+KEYS = ["Total", "Respiratory", "Coronavirus", "Influenza", "RSV", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
 df = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
+
+df = df.rename({"Syncytial Virus": "RSV"})
+
 raw_data = df.to_dict()
@@
 
-@fig-all_plot showcases that the total number of cases fluctuates with the number of respiratory cases, with number of outbreaks increasing since 2019 to 2022, and beginning to decrease afterwards.
-Additionally, the number of Coronavirus cases also fluctuates with the number of respiratory cases and total cases, following the same trend described above.
+@fig-all_plot showcases that the annual total number of outbreaks fluctuates with the annual number of respiratory outbreaks.
+The count of both of these has increased from 2019 to 2022, and began to decrease afterwards.
+Additionally, the number of Coronavirus outbreaks also fluctuates with the number of respiratory outbreaks and total outbreaks, following the same trend described above.
 
@@
 t = np.arange(2019, 2025, 1)
-KEYS = ["Influenza", "Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
+KEYS = ["Influenza", "RSV", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
 df = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
+df = df.rename({"Syncytial Virus": "RSV"})
+
 raw_data = df.to_dict()

### 2025-05-14 paper/paper.qmd
Fixed typo

@@
 
-abstract:  "Through the analysis of Outbreaks in Toronto Healthcare Institutions (2019-2025), this paper seeks to identify the trends in respiratory infections in Toronto healthcare institutions during the implementation and later lifting of COVID-19 preventative policies.
+abstract:  "Through the analysis of Outbreaks in Toronto Healthcare Institutions data (2019-2025), this paper seeks to identify the trends in respiratory infections in Toronto healthcare institutions during the implementation and later lifting of COVID-19 preventative policies.
 Through data visualization this analysis finds that while reported Coronavirus outbreaks have decreased since the pandemic, non-Coronavirus respiratory outbreaks such as Influenza, Parainfluenza (PIV), Respiratory Syncytial Virus (RSV), Metapneumovirus (MPV), and Rhinovirus (RV) outbreaks have risen from the lower levels they were at during the pandemic.

### 2025-05-15 scripts/02-download_data.py
added date to download, download to current year

@@ -12,6 +12,7 @@
 #### Workspace setup ####
 import polars as pl
 from io import StringIO
+from datetime import datetime
 
 #### Download data ####
 import requests
@@ -29,7 +30,8 @@ params = { "id": "outbreaks-in-toronto-healthcare-institutions"}
 package = requests.get(url, params = params).json()
 
 # To get resource data:
-i = 2025
+date = str(datetime.today()).split()[0]
+year = int(date[:4])
 for idx, resource in enumerate(package["result"]["resources"]):
 
        # for datastore_active resources:
@@ -40,5 +42,5 @@ for idx, resource in enumerate(package["result"]["resources"]):
 
             # Writing raw data to CSV
             analysis_data = pl.read_csv(StringIO(resource_dump_data.text))
-            analysis_data.write_csv("data/01-raw_data/ob_report_" + str(i) + ".csv")
-            i-=1
+            analysis_data.write_csv("data/01-raw_data/" + date + "_ob_report_" + str(year) + ".csv")
+            year-=1

### 2025-05-15 scripts/03-clean_data.py

@@ -10,8 +10,10 @@
 
 #### Workspace setup ####
 import polars as pl
+from datetime import datetime
 
 #constant variables
+date = str(datetime.today()).split()[0]
 MONTHS = ["January", "February", "March", "April", 
                              "May", "June", "July", "August", "September", "October", "November", "December"]
 YEARS = ["2019", "2020", "2021", "2022", "2023", "2024"]
@@ -38,11 +40,11 @@ yearly_disease_count["Total"] = [0, ]*6
 
 ## Finding composition based on primary causative agents ##
 
-for i in range(2019, 2025):
-    raw = pl.read_csv("data/01-raw_data/ob_report_" + str(i) + ".csv")
+for year in range(2019, 2025):
+    raw = pl.read_csv("data/01-raw_data/" + date + "_ob_report_" + str(year) + ".csv")
     for row in raw.rows(named = True):
        #index
-        index = i - 2019
+        index = year - 2019
 
        #getting disease type, only check common respiratory
         disease_type = row["Type of Outbreak"]
@@ -79,12 +81,12 @@ for i in range(2019, 2025):
                 disease_count["Other"][index] += 1
 
 ## summing total cases for the year
-for i in range(2016, 2026):
-    if i not in [2019, 2020, 2021, 2022, 2023, 2024]:
+for year in range(2016, 2026):
+    if year not in [2019, 2020, 2021, 2022, 2023, 2024]:
         continue
 
     # reading & iterating through csv
-    raw = pl.read_csv("data/01-raw_data/ob_report_" + str(i) + ".csv")
+    raw = pl.read_csv("data/01-raw_data/" + date + "_ob_report_" + str(year) + ".csv")
     for row in raw.rows(named = True):
         #getting disease type
         disease_type = row["Type of Outbreak"]
@@ -101,7 +103,7 @@ for i in range(2016, 2026):
             disease_two = row["Causative Agent - 2"]
 
         #updating yearly total count
-        index = i - 2019
+        index = year - 2019
         yearly_disease_count["Total"][index] += 1
 
         #updating yearly total respiratory count

### 2025-05-15 paper/paper.qmd
small fixes to code

@@
   - Evelyn Hughes
-thanks: "Code and data are available at: [https://github.com/EveHughes/Donaldson-Paper]."
+thanks: "Code and data are available at: https://github.com/EveHughes/Donaldson-Paper."
 date: 05/12/2025
@@
 import matplotlib.pyplot as plt
+from datetime import datetime
+
 pl.Config.set_tbl_hide_column_data_types(True)  
@@
 #reading in of data & removing unneeded info
-raw_data = pl.read_csv("../data/01-raw_data/ob_report_2019.csv")
+date = str(datetime.today()).split()[0]
+
+raw_data = pl.read_csv("../data/01-raw_data/" + date + "_ob_report_2019.csv")
 for key in raw_data.columns:
@@
 #rounding data
-summary_data = summary_data.with_columns(sc.by_dtype(pl.Float64).round(2))
+summary_data = summary_data.with_columns(sc.by_dtype(pl.Float64).cast(pl.Int64))
 
@@
 
-#rounding all data
-summary_data = summary_data.with_columns(sc.by_dtype(pl.Float64).round(2))
+#rounding all data & getting rid of quotes
+summary_data = summary_data.with_columns(sc.by_dtype(pl.Float64).cast(pl.Int64))
 

### 2025-05-15 paper/paper.qmd
added clarification to paper based on suggestions

@@
 
-According to the Ontario Ministry of Health, a respiratory outbreak is defined as “Two cases of acute respiratory infections (ARI) within 48 hours with any common epidemiological link (e.g., unit, floor), at least one of which must be laboratory-confirmed; OR Three cases of ARI (laboratory confirmation not necessary) occurring within 48 hours with any common epidemiological link (e.g., unit, floor).
-[@torontoPublicHealth2023].
+According to the Ontario Ministry of Health, a respiratory outbreak is defined as “Two cases of acute respiratory infections (ARI) within 48 hours with any common epidemiological link (e.g., unit, floor), at least one of which must be laboratory-confirmed; OR Three cases of ARI (laboratory confirmation not necessary) occurring within 48 hours with any common epidemiological link (e.g., unit, floor) [@torontoPublicHealth2023].
 
@@
 
-Causative Agent-1 is either a specific agent (ie: Coronavirus), or marked as "Unable to identify".
+Causative Agent-1 is either a specific agent (ie: Coronavirus), or marked as "Unable to identify". [^1]
+
+[^1]: Based on the definition of an outbreak, it is likely that Causative Agent-1 being marked as "Unable to identify" reflects an outbreak with three cases of ARI, for which laboratory confirmation is not necessary [@torontoPublicHealth2023].
+
 Causative Agent-2 is either a specific agent or marked as "".
@@
 Instances of Coronavirus and COVID-19 were grouped together under the category “Coronavirus”. Influenza A and Influenza B were additionally grouped together under the category “Influenza”. 
-@tbl-yearly_count_preview shows the data, and @tbl-yearly_count_summary shows the summary statistics.
+@tbl-yearly_count_preview [^2] shows the data, and @tbl-yearly_count_summary shows the summary statistics.
 
+[^2]: Totals between @tbl-yearly_count_preview and @tbl-disease_count_preview differ as the former aggregates the number of outbreaks, while the latter aggregates the number of non-null causative agents of which a given outbreak can have 0-2 of.
 ```{python}
@@
 raw_data = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
-raw_data = raw_data.head(5)
 raw_data = raw_data.rename({"Coronavirus": "Corona", "Syncytial Virus": "RSV",
@@
 While this could be representative of genuinely surpressed levels of other respiratory outbreaks due to increased COVID-19 cases, it could be indicative of a reporting issue.
-Further research should be conducted to clarify, though such research is beyond the scope of this paper.
+Additionally, in @tbl-yearly_count_preview, there is a near doubling of total cases between 2020 to 2024.
+The magnitude of this increase suggests that detection bias may have contributed.
+In other words, COVID-19 may have led to an increased focus on outbreak detection, and the increase we see from 2019 to 2024 was possibly caused in part by this.
+Further research should be conducted to clarify both of these issues, though such research is beyond the scope of this paper.
 
@@
 
-Additionally, as mentioned in [@sec-yearly_count], the extremely low reported outbreak counts of non-Coronavirus causative agents in 2021 point to potential issues with the validity of the data.
-Further research should be conducted to determine the cause of these low numbers and assess the validity of the dataset.
+Additionally, as mentioned in [@sec-yearly_count], the extremely low reported outbreak counts of non-Coronavirus causative agents in 2021 point to potential issues with the validity of the data, and the doubling in total outbreaks from 2020 to 2024 suggests detection biases may have contributed to pandemic and post-pandemic increases in outbreaks.
+Further research should be conducted to determine the cause of these low numbers and outbreak increases, and assess the validity of the dataset.
 
